\documentclass[11pt,a4paper]{article}
\usepackage[margin=2.2cm]{geometry}
\usepackage{parskip}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{array}
\usepackage{xcolor}
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection}{0.6em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.6em}{}
\newcommand{\up}{\textcolor{blue}{$\uparrow$}}
\newcommand{\dn}{\textcolor{red}{$\downarrow$}}

\title{Enrichment Analysis Results --- All Models\\[0.3em]
\large Modality labels vs.\ hallucination-driving neurons}
\author{MH-Neuron / Modality Taxonomy (step 10, halluc\_score)}
\date{June 13, 2026}

\begin{document}
\maketitle

\section*{Scope and status}
One section per model. Sections are regenerated from the enrichment JSONs on
disk by \texttt{build\_\allowbreak enrichment\_\allowbreak pdf.py}; re-running it picks up new combos as
their cluster jobs finish. \texttt{qwen2.5-vl-3b} is complete; the other models
are filled in incrementally and any combo not yet on disk is marked pending.

\textbf{What the numbers mean.} For each modality category (visual / text /
multimodal / unknown) we ablate the top 5\% of neurons by $\Delta$Hallucination
and run Fisher's exact test on whether that category is over-represented among
those hallucination-driving neurons. The \emph{odds ratio} (OR) is the headline:
OR $>1$ = enriched, OR $<1$ = depleted. Two tasks are scored: \textbf{POPE}
(visual object hallucination) and \textbf{TriviaQA} (text knowledge errors).


\section{Model: \texttt{qwen2.5-vl-3b}}
\emph{Complete --- all three label/hook combos.} All combos: top-$k$ = 5\%,
$n_{\text{driving}}=19{,}814$ of $396{,}288$ total neurons.

\subsection{POPE (visual task) --- odds ratios}
Hypothesis: \emph{visual} neurons should be enriched (ablating them blinds the
model). $p$-values in parentheses; $\approx$0 denotes $p<10^{-300}$.

\begin{center}
\begin{tabular}{lccc}
\toprule
category & PMBT $\cdot$ gate & PMBT $\cdot$ gate\_up & FT $\cdot$ gate \\
\midrule
visual     & \textbf{2.02} ($\approx$0)      & \textbf{1.38} (7e-106) & 1.18 (5e-25) \\
text       & 0.81 (3e-38)        & 0.91 (4e-9)            & 0.78 (1e-35) \\
multimodal & 1.04 (0.09, n.s.)   & 0.93 (8e-4)            & 0.93 (2e-5)  \\
unknown    & \textbf{0.03} ($\approx$0)      & \textbf{0.02} ($\approx$0)         & 1.06 (8e-5)  \\
\bottomrule
\end{tabular}
\end{center}

\subsection{TriviaQA (text task) --- odds ratios}
Hypothesis: \emph{text} neurons should be enriched (ablating them erases
knowledge). These were computed from the merged $\Delta H_{\text{TQA}}$ scores
(see Caveats).

\begin{center}
\begin{tabular}{lccc}
\toprule
category & PMBT $\cdot$ gate & PMBT $\cdot$ gate\_up & FT $\cdot$ gate \\
\midrule
visual     & 0.88 (3e-19)        & 1.40 (1e-115) & 1.48 (7e-140) \\
text       & \textbf{1.21} (1e-36)  & 0.86 (7e-23)  & 1.05 (0.01)   \\
multimodal & 0.85 (8e-11)        & 0.99 (0.6, n.s.) & 1.10 (1e-7) \\
unknown    & 1.03 (0.2, n.s.)    & 0.08 ($\approx$0) & 0.58 (1e-232) \\
\bottomrule
\end{tabular}
\end{center}

\subsection{The double dissociation (the key result)}
The dual-source design predicts a diagonal flip: visual neurons enriched on
POPE but \emph{not} TriviaQA, text neurons enriched on TriviaQA but \emph{not}
POPE. \up{} = enriched (OR$>$1), \dn{} = depleted (OR$<$1).

\begin{center}
\begin{tabular}{l cc cc}
\toprule
 & \multicolumn{2}{c}{\textbf{visual} OR} & \multicolumn{2}{c}{\textbf{text} OR} \\
combo & POPE & TQA & POPE & TQA \\
\midrule
\textbf{PMBT $\cdot$ gate}    & 2.02 \up & 0.88 \dn & 0.81 \dn & 1.21 \up \\
PMBT $\cdot$ gate\_up         & 1.38 \up & 1.40 \up & 0.91 \dn & 0.86 \dn \\
FT $\cdot$ gate               & 1.18 \up & 1.48 \up & 0.78 \dn & 1.05 \up \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Only PMBT $\cdot$ gate shows the clean flip} (visual \up\dn{},
text \dn\up). The other two do not: under PMBT gate\_up and FT gate, visual
neurons are enriched for \emph{both} tasks, so the modality axis is not
separated.

\subsection{Interpretation}

\textbf{Taxonomy: PMBT $\gg$ FT (same gate hook, POPE).}
PMBT concentrates visual function far harder (OR \textbf{2.02} vs FT's 1.18) and
\textbf{purges the ``unknown'' bucket to 0.03}, whereas FT leaves unknown
slightly \emph{enriched} at \textbf{1.06}. So FT's catch-all category still hides
hallucination-driving neurons that PMBT correctly reclassifies as visual. This is
the functional-validity argument for the permutation-test labels.

\textbf{Hook: gate $>$ gate\_up for separation.}
Both PMBT hooks share the qualitative signature (visual enriched, unknown
purged $\approx$0.02--0.03), confirming the effect under an \emph{independent}
labeling. But the \texttt{gate} hook gives sharper visual enrichment (2.02 vs
1.38) \emph{and} is the only one that yields the double dissociation; the
\texttt{gate\_up} intermediate captures visual function but does not separate
the text axis (visual enriched on TriviaQA too, OR 1.40).

\textbf{Headline.}
On identical neurons, ablation scores, and statistics, \textbf{PMBT labels at the
gate hook are the only configuration that (i) maximises visual enrichment, (ii)
empties the ``unknown'' bucket, and (iii) produces the visual/text double
dissociation across POPE and TriviaQA.} FT labels achieve none of these cleanly.

\subsection{Caveats}
\begin{itemize}
\item \textbf{TriviaQA enrichment was computed offline.} The 10C aggregation
historically wrote only the POPE \texttt{enrichment\_results.json}; the merged
$\Delta H_{\text{TQA}}$ existed (\texttt{ablation\_scores\_tqa.json}) but was not
run through enrichment. The TQA numbers here were produced by feeding that file
to the same \texttt{compute\_enrichment} routine. A pipeline fix (forwarding the
TQA scores so \texttt{enrichment\_results\_tqa.json} is emitted automatically)
has been applied for the remaining models.
\item Multimodal results are mostly non-significant or weak and are not load-bearing.
\item ``gate'' and ``gate\_up'' label \emph{different} neuron populations
(different hook point), so cross-hook agreement is corroboration by an
independent labeling, not a same-neuron comparison.
\end{itemize}

\section{Model: \texttt{internvl2.5-8b}}
\emph{Combos available: PMBT · gate: POPE=y, TQA=y; PMBT · gate\_up: POPE=y, TQA=y; FT · gate: POPE=--, TQA=--.}

top-$k$ = 5\%, $n_{\text{driving}}=22,937$ of 458,752 total neurons (first available combo).

\subsection{POPE (visual task) --- odds ratios}
Hypothesis: \emph{visual} neurons enriched. $p$ in parentheses; $\approx$0 denotes $p<10^{-300}$.
\begin{center}
\begin{tabular}{lcc}
\toprule
category & PMBT $\cdot$ gate & PMBT $\cdot$ gate\_up \\
\midrule
visual & 0.48 ($\approx$0) & \textbf{2.11} ($\approx$0) \\
text & \textbf{1.81} ($\approx$0) & 0.62 (3e-246) \\
multimodal & 0.98 (0.3, n.s.) & 0.71 (8e-98) \\
unknown & \textbf{22.82} (2e-15) & \textbf{0.00} (2e-18) \\
\bottomrule
\end{tabular}
\end{center}
\subsection{TriviaQA (text task) --- odds ratios}
Hypothesis: \emph{text} neurons enriched.
\begin{center}
\begin{tabular}{lcc}
\toprule
category & PMBT $\cdot$ gate & PMBT $\cdot$ gate\_up \\
\midrule
visual & \textbf{1.52} (5e-204) & 1.46 (5e-159) \\
text & 0.74 (9e-106) & 0.82 (6e-44) \\
multimodal & 0.83 (8e-24) & 0.82 (3e-37) \\
unknown & 0.00 (0.4, n.s.) & 0.21 (3e-09) \\
\bottomrule
\end{tabular}
\end{center}
\subsection{Double dissociation}
\up{} = enriched (OR$>$1), \dn{} = depleted (OR$<$1). Only combos with both tasks scored are shown.
\begin{center}
\begin{tabular}{l cc cc}
\toprule
 & \multicolumn{2}{c}{\textbf{visual} OR} & \multicolumn{2}{c}{\textbf{text} OR} \\
combo & POPE & TQA & POPE & TQA \\
\midrule
PMBT $\cdot$ gate & 0.48 \dn & 1.52 \up & 1.81 \up & 0.74 \dn \\
PMBT $\cdot$ gate\_up & 2.11 \up & 1.46 \up & 0.62 \dn & 0.82 \dn \\
\bottomrule
\end{tabular}
\end{center}
\section{Model: \texttt{llava-onevision-7b}}
\emph{Combos available: PMBT · gate: POPE=--, TQA=--; PMBT · gate\_up: POPE=--, TQA=--; FT · gate: POPE=--, TQA=--.}

\emph{No enrichment results on disk yet --- layers still running.}
\section{Model: \texttt{llava-1.5-7b}}
\emph{Combos available: PMBT · gate: POPE=--, TQA=--; PMBT · gate\_up: POPE=--, TQA=--; FT · gate: POPE=--, TQA=--.}

\emph{No enrichment results on disk yet --- layers still running.}

\end{document}
